---
title: "Simulation-Based Validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation-Based Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(xpar)
library(dplyr)
```

## Why simulate

A real-data case study can show that `xpar` recovered the correct parents
for one dataset. It cannot show *how often* the method would get the right
answer under conditions you have not observed yet, or *where* it starts to
fail. That requires data with a known, controlled ground truth — which real
pedigrees do not give you, and simulated interspecific crosses do.

Every result in this vignette is generated by [`simulate_hybrid_pedigree()`],
which uses `AlphaSimR::runMacs(split = ...)` to create two founder pools
that diverged a controllable number of generations ago, crosses one
individual from each to produce a known F1 hybrid, and leaves the rest as
decoy candidates. [`inject_noise()`] then overlays a controllable amount of
genotyping error and missingness, and [`run_trial()`] / [`benchmark_grid()`]
score whether [`block_bootstrap()`] recovered the true pair. The cached
results below (`sim_study`) come from `data-raw/simulation_study.R`, fully
reproducible from its recorded seeds.

```{r}
data(sim_study)
names(sim_study)
```

All four sweeps share a baseline: divergence of 2000 generations between
the two simulated species, 10 candidates per species (190 candidate pairs),
10 linkage groups, and 200 segregating sites per chromosome. Each sweep
varies one axis at a time.

## 1. Genotyping error rate

```{r}
sim_study$error_broad |>
  group_by(divergence, error_rate) |>
  summarise(
    power_top1 = mean(true_pair_is_top),
    mean_win_prob = mean(true_pair_win_prob),
    .groups = "drop"
  )
```

Recovery of the true pair as the single best-ranked candidate stays at
100% through 10% injected genotyping error at both low (20-generation) and
high (2000-generation) divergence, and holds to 25% at low divergence. At
high divergence it starts to erode earlier — 80% power by 25% error — which
the next section explains. In every case, win probability moves first: at
high divergence it has already fallen to 0.71 by 25% error, well before the
80%-power point it foreshadows.

A finer sweep around high divergence localizes exactly where recovery
breaks down:

```{r}
sim_study$error_fine |>
  group_by(error_rate) |>
  summarise(
    power_top1 = mean(true_pair_is_top),
    mean_win_prob = mean(true_pair_win_prob),
    mean_rank = mean(true_pair_rank),
    .groups = "drop"
  )
```

The transition is a zone (roughly 30-38% error), not a cliff: power falls
from 90% at 25% error to 30% by 35-38% error, while win probability has
already dropped below 0.5 by 30% error — again, well ahead of the rank
collapse. (This finer sweep uses its own independent replicates from the
broad sweep above, so the two do not agree to the percentage point at the
shared 25% condition — both are noisy estimates from a handful of
replicates, not a contradiction.)

**Why high divergence is not simply "easier."** A naive expectation is that
more interspecific divergence should always help, since it gives the true
pair a more distinctive signature. It does — up to a point. But higher
divergence also increases the fraction of loci that are near-fixed
differences between the two species (parent genotypes like homozygous-ref x
homozygous-alt), and those loci have the *least* tolerance for genotyping
error: there is exactly one valid offspring genotype, versus loci with a
heterozygous parent, which tolerate any offspring genotype. So heavy
genotyping noise erodes the signal from highly diverged crosses faster than
from weakly diverged ones, not slower. This is a real, testable property of
the method, not an artifact — and it is the kind of failure mode a
fixed-error-rate tool has no mechanism to reveal, because it never asks the
question in the first place.

## 2. Missingness

```{r}
sim_study$missingness |>
  group_by(missing_rate) |>
  summarise(
    power_top1 = mean(true_pair_is_top),
    mean_win_prob = mean(true_pair_win_prob),
    .groups = "drop"
  )
```

At a fixed, moderate 20% genotyping error rate, missingness alone is
gentler than error: power holds at 100% through 20% missing calls, and only
degrades (to 87.5%, then 75%) once missingness exceeds 30% on top of that
error rate. Win probability again declines smoothly and well before power
does, from 0.99 at 0% missing to 0.62 at 30%.

## 3. Candidate-pool size

```{r}
sim_study$pool_size |>
  group_by(n_per_species) |>
  summarise(
    n_pairs = mean(n_pairs),
    power_top1 = mean(true_pair_is_top),
    mean_win_prob = mean(true_pair_win_prob),
    .groups = "drop"
  )
```

Rank recovery is essentially unaffected by how many decoy candidates are in
the pool — the true pair is the point-estimate best in every replicate from
3 candidates per species (15 pairs) up to 10 (190 pairs), and in 87.5% of
replicates even at 20 per species (780 pairs). Win probability moves much
more than power does: it falls steadily from 0.996 to 0.72 as the pool
grows. This is exactly the behavior a
method should have when testing more candidate pairs increases the chance
that some decoy wins a given bootstrap replicate by chance alone — the
ranking stays correct, but the *confidence* attached to it correctly
reflects the larger multiple-comparison space, the same logic behind the
multiple-testing caveat already carried in the single-candidate report
for the real GNT43 dataset.

## Summary

Across three largely independent stress axes — genotyping error,
missingness, and candidate-pool size — `xpar` shows the same qualitative
pattern: point-estimate rank recovery is robust well past the point where a
naive user might expect trouble, and where it does eventually fail, the
bootstrap win probability gives advance warning by declining smoothly
first. That graceful-degradation property, not just a headline recovery
rate, is the result worth reporting alongside the real-data GNT43 case
study.
